A weakly structured stem for human origins in Africa

### 2023-03-15 paper.tex

We also find that Stem 2 continued to contribute to western Africans during the
Last Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an
interpretation is reinforced by the differential migration rates between
regions, i.e., the gene flow from Stem 2 to western Africans is estimated to be
five times that of the rate to eastern Africans during this period. We
performed a variety of validation tests to explore sensitivity of our
assumptions, including relaxing fixed parameters (SI
Section~\ref{sec:validation_strategy}). Most validation tests resulted in
parameters similar to the models discussed above. However, one exception was
the inferred Out-of-Africa and eastern/western African divergences which were
10--15ka younger than our fixed parameters. These younger dates are at odds
with the accepted timing of the Out-of-Africa expansion that contributed to
later human populations at approximately $\>50$ka, as based on archaeological,
climatic and fossil information \citep{Fu2014-am,Groucutt2015-do,
Prufer2021-ao,Beyer2021-ym}. We interpret the free estimate for eastern African
vs. European divergence as reflecting our inclusion of only a single
Out-of-Africa population in the model, the lack of a nearby source for
back-to-Africa gene flow, and other regionally complex parameters. However,
older pan-African features of our inferred models are minimally affected by
this analysis choice (SI Section~\ref{sec:models}).

>>>>

We also find that Stem 2 continued to contribute to western Africans during the
Last Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an
interpretation is reinforced by the differential migration rates between
regions, i.e., the gene flow from Stem 2 to western Africans is estimated to be
five times that of the rate to eastern Africans during this period. We
performed a variety of validation tests to explore sensitivity of our
assumptions, including relaxing fixed parameters (SI
Section~\ref{sec:validation_strategy}). Most validation tests resulted in
parameters similar to the models discussed above.  However, one exception was
the inferred Out-of-Africa and eastern/western African divergences which were
10–15ka younger than our fixed parameters. These younger dates are at odds with
the accepted timing of the Out-of-Africa expansion that contributed to later
human populations at approximately 50ka, as based on archaeological, climatic
and fossil information \citep{Fu2014-am,Groucutt2015-do,
Prufer2021-ao,Beyer2021-ym}. Because the inference approach is unbiased in
simulations, we interpret the free estimate for eastern African vs. European
divergence as reflecting our inclusion of only a single Out-of-Africa
population in the model, the lack of a nearby source for back-to-Africa gene
flow, and other regionally complex parameters, rather than to a systematic bias
that may affect all parameters in the model. Older pan-African features of our
inferred models are minimally affected by the choice of these fixed parameters
(SI Section~\ref{sec:change-dates}).
